Hydra battles in Coq

### 2020-10-25 doc/hydras.tex
Document gamma0 (#5)
* initialization of a documentation on Gamma0; strongly critical ordinals

* redaction of chapter on Gamma0 (partial)

* Writing a chapter on Gamma0

* First draft of the chapter on Gamma0 (to be reviewed).

* Corrections to Gamma0 (chapter and scripts)

* First draft of the chapter on Gamma0

The ordinal $\omega^2$ (also called $\varphi_0(2)$, see Chap.~\ref{chap:schutte}), is represented by the lexicographically ordered cartesian product $\mathbb{N}\times\mathbb{N}$.

@@

Note that if the notation $N_A$ implements the ordinal 
$\alpha$,  the new notation $\omega^{N_A}$ must implement the ordinal $\varphi_0(\alpha)$, a.k.a. $\omega^\alpha$ (see chapter~\ref{chap:schutte})

@@

\subsubsection{The ordinal \(\omega^\alpha\), a.k.a. \(\varphi_0(\alpha)\)}
\label{sec:orgheadline70}

@@

\begin{remark}
\label{sec:orgheadline69}
The name \(\varphi_0\)
   comes from ordinal numbers theory. In~\cite{schutte}, Schütte defines 
$\varphi_0$  as the ordering (\emph{i.e.} enumerating) function of the set  of \emph{additive principal ordinals} \emph{i.e.} strictly positive ordinals $\alpha$ that verify $\forall \beta<\alpha, \beta+\alpha=\alpha$. For Schütte,  $\omega^\alpha$ is just a notation for $\varphi_0(\alpha)$.  See also Chapter~\ref{chap:schutte} of this document.
\end{remark}

@@

\subsection{Pretty printing Ordinals in Cantor normal Form}

@@

@@

We would like to get exact formulas for the ordinal $\omega^2$, a.k.a.
$\varphi_0(2)$. This ordinal is the limit of the sequence $\omega\times i\;(i \in \mathbb{N}$. Thus, we have to study ordinals of this form, then use 
our lemma on limits.

@@

In this section, we define the function which maps any $\alpha\in\mathbb{O}$ to
the ordinal  $\omega^\alpha$, also written 
$\varphi_0(\alpha)$. 
It is an opportunity to apply the definitions and results of the preceding section. 
Indeed,  Schütte first defines a subset of $\mathbb{O}$: the set of additive principal ordinals, and $\varphi_0$  is just defined as the ordering function of this set.

@@

Let us call  $\varphi_0$ the ordering function of \texttt{AP}.
In the mathematical text, we shall use indifferently the notations  $\omega^\alpha$ and$\varphi_0(\alpha)$.

@@

Using $\varphi_0$, we can define recursively the set of finite omega-towers.

@@

The rest of our library \texttt{AP} is devoted to the proof of properties of additive principal ordinals, hence of the ordering function  $\varphi0$ and the ordinal $\epsilon_0$ (which we could not express within the type \texttt{T1}).

@@

\subsubsection{Properties of the function \texorpdfstring{$\varphi_0$}{phi0}}

The ordering function $\varphi_0$ of the set \texttt{AP} is defined on the full set $\mathbb{O}$ and is continuous (Schütte says that this function is \emph{normal}).

@@

The following properties come from  the definition of $\varphi_0$ as the ordering function of \texttt{AP}. It may be interesting to compare these proofs with the computational ones described in Chapter ~\ref{chap:T1}.

@@

Please note that the lemma \texttt{omega\_eq} above, is consistent with the interpretation of the ordering function $\varphi_0$ as the exponential of basis $\omega$. Indeed we could have written this lemma with our alternative notation:

@@

The proof is as follows:
\begin{enumerate}
\item Since $\alpha<\epsilon_0$, consider the least $i$ such that $\alpha$ is strictly less than the omega-tower of height $i$.
\item
  \begin{itemize}
  \item If $i=0$, then the result is trivial (because $\alpha=0$)
 \item  Otherwise let $i=j+1$; 
          $\alpha$ is greater than or equal to the omega-tower of height $j$.
         By monotonicity,  $\varphi_0(\alpha)$ is greater than or equal to 
        the omega-tower of height $j+1$, thus strictly greater than $\alpha$
  \end{itemize}
 \end{enumerate}

Moreover,  $\epsilon_0$ is the least ordinal $\alpha$ that verifies the equality 
$\alpha = \omega^\alpha$, in other words the least fixpoint of the function  $\varphi_0$.

@@

The ordinal $\Gamma_0$ is defined in Section 13 of ~\cite{schutte} as the least \emph{strongly critical ordinal}. It is widely known as the \emph{Feferman-Schütte ordinal}.
Section V, 13 of~\cite{schutte} defines \emph{strongly critical} and
\emph{maximal $\alpha$-critical} ordinals.

@@

Definition maximal_critical alpha : Ensemble Ord :=
  fun gamma =>
    In (Cr alpha) gamma /\
    forall ksi, alpha < ksi -> ~ In (Cr alpha) ksi.
\end{Coqsrc}

\begin{project}
In order to define $\Gamma_0$ as a n inhabitant of type \texttt{Ord}, you should prove that the set of strongly critical ordinal is not empty, hence has a least element.
\end{project}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=11cm]{epsilon0.jpg}
  \caption{Discrete mathematics and formal proofs}
  \label{fig:gamma0}
\end{figure}

@@

Like in chapter~\ref{chap:T1}, we get familiar with the type \texttt{T2} by recognising simple constructs like finite ordinals, $\omega$, etc., as inhabitants of ÷texttt{T2}.

@@

Notation "'FS' n" := (gcons zero zero n zero) (at level 10) : T2_scope.

(** the [n]-th ordinal  *)

Definition fin (n:nat) := match n with 0 => zero | S p => FS p end.

@@

We are also able to give a term to $\epsilon_0$, and even to generalize it!.

@@

The justification of the definition of \texttt{epsilon0} will be given in further sections.

@@

\subsection{Examples}
Seven constructors! In order to get accustomed with this definition, let us look at a small set of examples, covering all the constructors of \texttt{lt}.

@@

Fixpoint t2_length (t:T2) : nat :=
  match t  with 
    zero => 0
  | gcons a b n v => 
       nbterms (gcons a b n v) + 
      2 * (Max.max (t2_length a)
                              (Max.max (t2_length b) 
                                                (t2_length_aux v)))
  end
with t2_length_aux (t:T2) : nat :=
 match t with zero => 0
            | gcons a b n v =>
               Max.max (t2_length a) 
                                (Max.max (t2_length b) (t2_length_aux v))
 end.
\end{Coqsrc}

@@

\subsection{The Veblen function \texorpdfstring{$\varphi$}{\texttt{phi}}}

@@

For instance, the following theorem states that, if $\gamma$ is the sum of a limit ordinal $\beta$ and a finite ordinal $n$, and $\beta$ is a fixpoint of
$\varphi(\alpha)$, then $\psi(\alpha,\gamma)=\phi(\alpha, \gamma+1)$.

@@

On the other hand, $\varphi$ can be expressed in terms of $\psi$.

@@

@@

Prove that the implemantations of \texttt{succ}, \texttt{+}, $\varphi_0$, etc.
are compatible in both notation systems.

Note that a function \texttt{T1\_inj} from \texttt{T1} to \texttt{T2} has already been defined. It may help to complete the task.
\end{project}

@@

\begin{table}[h]
  \centering
  \begin{threeparttable}
    \caption{Other notations}
\begin{tabular}{|c|c|c|c|l|}
\hline
Name & Gallina&Math& Description& Page \\\hline
F & \texttt{F $n$} & $n$ & The $n$-th finite ordinal &  
\pageref{sect:notation-F}, \pageref{sect:notation-F-sch}\\ 
FS & \texttt{FS $n$} & $n+1$ & The $n+1$-th finite ordinal  \tnote{2} &  
\pageref{sect:notation-FS}\\ 
omega & \texttt{omega} & $\omega$ &   the first infinite ordinal   & \pageref{sect:notation-omega}, \pageref{sect:omega-T1}, \pageref{sect:omega-notation2}, \dots\\
phi0     & \texttt{phi0 alpha} & $\varphi_0(\alpha),\; \omega^\alpha$&exponential of base $\omega$ & \pageref{sect:notation-phi0}\\

>>>>

The ordinal $\omega^2$ (also called $\phi_0(2)$, see Chap.~\ref{chap:schutte}), is represented by the lexicographically ordered cartesian product $\mathbb{N}\times\mathbb{N}$.

@@

Note that if the notation $N_A$ implements the ordinal 
$\alpha$,  the new notation $\omega^{N_A}$ must implement the ordinal $\phi_0(\alpha)$, a.k.a. $\omega^\alpha$ (see chapter~\ref{chap:schutte})

@@

\subsubsection{The ordinal \(\omega^\alpha\), a.k.a. \(\phi_0(\alpha)\)}
\label{sec:orgheadline70}

@@

\begin{remark}
\label{sec:orgheadline69}
The name \(\phi_0\)
   comes from ordinal numbers theory. In~\cite{schutte}, Schütte defines 
$\phi_0$  as the ordering (\emph{i.e.} enumerating) function of the set  of \emph{additive principal ordinals} \emph{i.e.} strictly positive ordinals $\alpha$ that verify $\forall \beta<\alpha, \beta+\alpha=\alpha$. For Schütte,  $\omega^\alpha$ is just a notation for $\phi_0(\alpha)$.  See also Chapter~\ref{chap:schutte} of this document.
\end{remark}

@@

\subsection{Pretty printing Ordinals in Cantor normal Form}
\label{sect:ppT1}

@@

\index{Projects}
\begin{project}
Design  (in \ocaml?) a set of tools for systematically pretty printing ordinal terms in Cantor normal form.
\end{project}

@@

We would like to get exact formulas for the ordinal $\omega^2$, a.k.a.
$\phi_0(2)$. This ordinal is the limit of the sequence $\omega\times i\;(i \in \mathbb{N}$. Thus, we have to study ordinals of this form, then use 
our lemma on limits.

@@

In this section, we define the function which maps any $\alpha\in\mathbb{O}$ to
the ordinal  $\omega^\alpha$, also written 
$\phi_0(\alpha)$. 
It is an opportunity to apply the definitions and results of the preceding section. 
Indeed,  Schütte first defines a subset of $\mathbb{O}$: the set of additive principal ordinals, and $\phi_0$  is just defined as the ordering function of this set.

@@

Let us call  $\phi_0$ the ordering function of \texttt{AP}.
In the mathematical text, we shall use indifferently the notations  $\omega^\alpha$ and$\phi_0(\alpha)$.

@@

Using $\phi_0$, we can define recursively the set of finite omega-towers.

@@

The rest of our library \texttt{AP} is devoted to the proof of properties of additive principal ordinals, hence of the ordering function  $\phi0$ and the ordinal $\epsilon_0$ (which we could not express within the type \texttt{T1}).

@@

\subsubsection{Properties of the function \texorpdfstring{$\phi_0$}{phi0}}

The ordering function $\phi_0$ of the set \texttt{AP} is defined on the full set $\mathbb{O}$ and is continuous (Schütte says that this function is \emph{normal}).

@@

The following properties come from  the definition of $\phi_0$ as the ordering function of \texttt{AP}. It may be interesting to compare these proofs with the computational ones described in Chapter ~\ref{chap:T1}.

@@

Please note that the lemma \texttt{omega\_eq} above, is consistent with the interpretation of the ordering function $\phi_0$ as the exponential of basis $\omega$. Indeed we could have written this lemma with our alternative notation:

@@

The proof is as follows:
\begin{enumerate}
\item Since $\alpha<\epsilon_0$, consider the least $i$ such that $\alpha$ is strictly less than the omega-tower of height $i$.
\item
  \begin{itemize}
  \item If $i=0$, then the result is trivial (because $\alpha=0$)
 \item  Otherwise let $i=j+1$; 
          $\alpha$ is greater than or equal to the omega-tower of height $j$.
         By monotonicity,  $\phi_0(\alpha)$ is greater than or equal to 
        the omega-tower of height $j+1$, thus strictly greater than $\alpha$
  \end{itemize}
 \end{enumerate}

Moreover,  $\epsilon_0$ is the least ordinal $\alpha$ that verifies the equality 
$\alpha = \omega^\alpha$, in other words the least fixpoint of the function  $\phi_0$.

@@

The ordinal $\Gamma_0$ is defined in Section 13 of ~\cite{schutte} as the least \emph{strongly critical ordinal}. It is widely known as the \emph{Feferman-Schütte ordinal}.

Section V, 13 of~\cite{schutte} defines \emph{strongly critical} and
\emph{maximal $\alpha$-critical} ordinals.

We give their definition in \coq's syntax.

@@

Definition maximal_critical alpha : Ensemble Ord :=
  fun gamma =>
    In (Cr alpha) gamma /\
    forall ksi, alpha < ksi -> ~ In (Cr alpha) ksi.

Definition Gamma0 := the_least strongly_critical.
\end{Coqsrc}

\index{Projects}
\begin{project}
Prove that a (countable)  ordinal $\alpha$ is strongly critical iff $\phi \alpha 0=\alpha$ (Theorem 13.13 of~\cite{schutte} ). 
\end{project}

\index{Projects}
\begin{project}
Prove that the set of strongly critical ordinals is unbounded and closed (Theorem 13.14 of~\cite{schutte} ). Thus this set is not empty,  hence has a least element. Otherwise, the definition of $\Gamma_0$ above would be useless.
\end{project}

@@

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=11cm]{epsilon0.jpg}
  \caption{Veblen normal form}
  \label{fig:gamma0}
\end{figure}

Like in chapter~\ref{chap:T1}, we get familiar with the type \texttt{T2} by recognising simple constructs like finite ordinals, $\omega$, etc., as inhabitants of \texttt{T2}.

@@

(** The (n+1)-th finite ordinal *)
Notation "'FS' n" := (gcons zero zero n zero) (at level 10) : T2_scope.

(** the [n]-th ordinal  *)
Definition fin (n:nat) := match n with 0 => zero | S p => FS p end.

@@

@@

@@

Seven constructors! In order to get accustomed with this definition, let us look at a small set of examples, covering all the constructors of \texttt{lt}.

\subsection{Examples}

@@

Fixpoint t2_length (t:T2) : nat :=
  match t  with 
    zero => 0
  | gcons a b n v => 
       nbterms (gcons a b n v) + 
      2 * (Max.max (t2_length a)
                              (Max.max (t2_length b) 
                                                (t2_length_aux v)))
  end
with t2_length_aux (t:T2) : nat :=
 match t with 
 | zero => 0
  | gcons a b n v =>
           Max.max (t2_length a) 
                            (Max.max (t2_length b) (t2_length_aux v))
 end.
\end{Coqsrc}

@@

\subsection{The Veblen function \texorpdfstring{$\phi$}{\texttt{phi}}}

@@

For instance, the following theorem states that, if $\gamma$ is the sum of a limit ordinal $\beta$ and a finite ordinal $n$, and $\beta$ is a fixpoint of
$\phi(\alpha)$, then $\psi(\alpha,\gamma)=\phi_\alpha(\gamma+1)$.

@@

On the other hand, $\phi$ can be expressed in terms of $\psi$.

@@

\index{Projects}
\begin{project}
Please study a way to pretty print ordinal terms in Veblen normal form (see Section~\vref{sect:ppT1}).
\end{project}

@@

Prove that the implemantations of \texttt{succ}, \texttt{+}, $\phi_0$, etc.
are compatible in both notation systems.

Note that a function \texttt{T1\_inj} from \texttt{T1} to \texttt{T2} has already been defined. It may help to complete the task.

\noindent\emph{From \href{../src/html/hydras.Gamma0.T2.html\#T1_to_T2}%
{\texttt{Gamma0.T2}}}
\begin{Coqsrc}
(* injection from T1 *)

Fixpoint T1_to_T2 (alpha :T1) : T2 :=
  match alpha  with
  | T1.zero => zero
  | T1.ocons a n b => gcons zero (T1_to_T2 a) n (T1_to_T2 b)
  end.
\end{Coqsrc}

\begin{project}
Prove that the notation system \texttt{Gamma0} is a correct implementation 
of the segment $[0,\Gamma_0)$ of the set of countable ordinals.
\end{project}

@@

\begin{table}[h]
  \centering
  \begin{threeparttable}
    \caption{Other notations}
\begin{tabular}{|c|c|c|c|l|}
\hline
Name & Gallina&Math& Description& Page \\\hline
F & \texttt{F $n$} & $n$ & The $n$-th finite ordinal &  
\pageref{sect:notation-F}, \pageref{sect:notation-F-sch}\\ 
FS & \texttt{FS $n$} & $n+1$ & The $n+1$-th finite ordinal  \tnote{2} &  
\pageref{sect:notation-FS}\\ 
omega & \texttt{omega} & $\omega$ &   the first infinite ordinal   & \pageref{sect:notation-omega}, \pageref{sect:omega-T1}, \pageref{sect:omega-notation2}, \dots\\
phi0     & \texttt{phi0 alpha} & $\phi_0(\alpha),\; \omega^\alpha$&exponential of base $\omega$ & \pageref{sect:notation-phi0}\\
